Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 overview.tex
Removed commented-out text from lots of files - basically vestigial junk.

%\subsubsection{Magnetohydrodynamics and Gravity}

@@

\begin{eqnarray} 
  \frac{\partial \rho}{\partial t} 
  + \frac{1}{a} \div (\rho \vecv) & = & 0,
  \label{eq:mass} \\
%
  \frac{\partial \rho \vecv}{\partial t}  
  + \frac{1}{a} \div \left(\rho \vecv \vecv + \myvec{I}p^* - 
    \frac{\vecB \vecB}{a} \right) & = &
  - \frac{\dot{a}}{a} \rho \vecv - \frac{1}{a} \rho \grad \phi,
% 
%          + \frac{1}{a} ( \vecv \cdot \grad ) \vecv
%    & = & - \frac{\dot{a}}{a} \vecv 
%          - \frac{1}{a \rho} \grad p
%          - \frac{1}{a \rho} (\nalba \times \vecB) \times \vecB
%          - \frac{1}{a} \grad \phi, 
  \label{eq:momentum} \\
%
  \frac{\partial E}{\partial t} 
  + \frac{1}{a} \div \left[ (E + p^*) \vecv - 
    \frac{1}{a} \vecB (\vecB \cdot \vecv) \right] & = &
  - \frac{\dot{a}}{a} \left( 2E - \frac{B^2}{2a} \right) - 
  \frac{\rho}{a} \vecv \cdot \grad \phi 
  - \Lambda + \Gamma + \frac{1}{a^2} \div \fcond,
% 
%         + \frac{1}{a} \vecv \cdot \grad E
%  & = & - \frac{\dot{a}}{a} (3 \frac{p}{\rho} + {\vecv \cdot \vecv} )
%       - \frac{1}{a \rho } \div (p \vecv) 
%%\nonumber   & &   %uncomment for 2-column
%        - \frac{1}{a} \vecv \cdot \grad \phi 
%               + \Gamma - \Lambda.
  \label{eq:total_energy}  \\
%
  \frac{\partial \vecB}{\partial t} - 
  \frac{1}{a}  \grad \times (\vecv \times \vecB) & = & 
  %-\frac{\dot{a}}{2 a} \vecB. 
  0
  \label{eq:induction}
\end{eqnarray}

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \grad \rho
%    & = & - \rho \div \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \grad ) \vecv 
%    & = & - \frac{1}{\rho} \grad p
%          - \grad \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \grad E
%    & = & - \frac{1}{\rho} \div (p \vecv) 
%          - \vecv \cdot \grad \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
In these equations, $E$, $\rho$, $\vecv$, and $\vecB$ are the comoving
total fluid energy density, comoving gas density, peculiar
velocity, and comoving magnetic field strength, respectively. The
matrix 
$\myvec{I}$ is the identity matrix, and $a$ is the cosmological
expansion parameter (discussed in more detail below).  The first
equation represents conservation of mass, the second conservation of
momentum, and the third conservation of total (kinetic plus thermal
plus magnetic) fluid energy.  They are respectively, the first, second, and third
moments of the Boltzmann equation.  The fourth equation is the
magnetic induction equation.  Terms representing radiative cooling
($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the
energy equation (\ref{eq:total_energy}), as does the flux due to
thermal heat conduction ($\fcond$).

@@

\begin{eqnarray}
\vecx & = & \vecx' / a, \\
\vecv & = & a \hspace{0.5mm} \dot{\vecx} = 
              \vecvp - \dot{a}\vecx, \\
\rho  & = & a^3 \rhop,   \\
p     & = & a^3 \pp, \\
E     & = & a^3 \left(\Ep - 
            \dot{a} \, \vecx \cdot \vecvp - 
            \frac{1}{2} \dot{a}^2 \vecx^2\right), \\
\phi  & = & \phip + \frac{1}{2} \, a \ddot{a} \, \vecx^2, \\
\vecB & = & a^2 \vecB^\prime,
\end{eqnarray}
where primes indicate quantities in proper coordinates (fixed frame).
We note that the definition of comoving $\vecB$ used here is such
that a uniform field is constant in a homogeneous expanding universe
and is also the quantity used in the MHD-CT solver described
below; however, it is not universal,
and is slightly different from that used both in \citet{Li08a}
and in the MHD-Dedner solver (see section~\ref{sec.num.hydro-muscl}
for more details).
%These definitions result in the following comoving equivalents to eqs. 
%(\ref{eq:mass})-(\ref{eq:potential}).
%%
%
%\begin{eqnarray} 
%\frac{\partial \rhop}{\partial t} 
%          + \frac{1}{a} \vecvp \cdot \grad \rhop 
%    & = & - \frac{1}{a} \rhop \div \vecvp, 
%        \label{eq:mass_prime} \\
%%
%\frac{\partial \vecvp}{\partial t}  
%          + \frac{1}{a} ( \vecvp \cdot \grad ) \vecvp 
%    & = & - \frac{\dot{a}}{a} \vecvp 
%          - \frac{1}{a \rhop} \grad \pp 
%          - \frac{1}{a} \grad \phip, 
%        \label{eq:momentum_prime} \\
%%
%\frac{\partial \Ep}{\partial t} 
%         + \frac{1}{a} \vecvp \cdot \grad \Ep 
%   & = & - \frac{\dot{a}}{a} (3 \frac{\pp}{\rhop} + {\vecvp \cdot \vecvp} )
%         - \frac{1}{a \rhop } \div (p \vecvp) 
%%\nonumber   & &   %uncomment for 2-column
%         - \frac{1}{a} \vecvp \cdot \grad \phip 
%                + \Gamma - \Lambda.
%       \label{eq:total_energy_prime} \\
%%
%\Ep  & = & \ep + \frac{1}{2} {\vecvp \cdot \vecvp},
%        \label{eq:total_energy_def_prime} \\
%%
%\ep                  & = & 
%        \pp/\left[ \left( \gamma - 1 \right) \rhop \right],
%      \label{eq:state_prime} \\ 
%{\nabla}^2 \phip    & = & 
%        \frac{4\pi G}{a} ( \rhop_{total} - \rhop_0 ).
%      \label{eq:poisson}
%\end{eqnarray}
%%
%%We use the proper peculiar baryonic velocity $\vecv_b \equiv a(t) d\vecx / dt$, proper pressure $p$, and modified gravitational potential $ \phi $ which is related to the potential in proper coordinates $\Phi$ by 
%%$$\phi \equiv \Phi + \frac{1}{2} a \ddot{a} \vecx^2. $$
%%The density, however, is comoving:
%%$$\rho_b \equiv \rho_{b,proper}a(t)^3, $$
The expansion parameter $a \equiv 1/(1 + z)$ follows the expansion of
a smooth, homogeneous background, where $z$, the redshift, is a
function only of $t$.  All spatial derivatives are determined with
respect to the comoving position $\vecx$, which removes the universal
expansion from the coordinate system. The evolution of $a(t)$ is
governed by the second Friedmann equation for the expansion of a
spatially homogeneous and isotropic universe
%
\begin{eqnarray}
\frac{\ddot{a}}{a} & = & 
      - \frac{4 \pi G }{3 a^3 } (\rho_0 
      + 3p_0/c^2) 
      + \Lambda_c c^2 / 3 .
      \label{eq:expansion} 
\end{eqnarray}
%
Here $\rho_0$ is the mean comoving mass density (including both
baryonic and dark matter), $p_0$ is the comoving background pressure
contribution, and $\Lambda_c$ is the cosmological constant.
%$\gamma$ is the ratio of specific heats, and $\Lambda$ is the cosmological constant.
%We note that if $\gamma$ is not $5/3$ then $E$ in the first term on
%the right side of eq.~(ref{eq:total_energy}) must be replaced with 
%eq.~(\ref{eq:state}).
This system of equations is limited to the non-relativistic regime and
assumes that curvature effects are not important --- both assumptions
are reasonable as long as the size of the simulated region is small
compared to the radius of curvature and the Hubble length $c/H$, where
$c$ is the speed of light and $H = \dot{a}/a$ is the Hubble constant.

@@

%\subsubsection{Particles}

@@

%\subsubsection{Chemistry}

@@

%\subsubsection{Radiation transport: moment method}

As an alternative to the ray-casting strategy, \enzo\ can also solve
the angle-averaged radiative transfer equation through a
flux-limited diffusion approximation, with couplings to both the gas
energy and chemical number densities. The equations solved are
\begin{eqnarray}
  \label{eq:fld_radiation}
  \frac{\partial E_r}{\partial t} + \frac1a \div\left(E_r \vecv \right) &=& 
  \div\left(D\grad E_r\right) -
  \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
  \label{eq:fld_heating}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c + \Gamma - \Lambda,
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \div\left({\mathrm n}_i \vecv_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
\end{eqnarray}
where $E_r = E_r(\vecx,t)$ is a grey radiation energy density and
$e_c$ is the internal energy correction due to photo-heating and
chemical cooling.
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  
Here, we define $E_r$ through first assuming a fixed frequency
spectrum, i.e.  $E_{\nu}(\nu,\vecx,t) = \tilde{E}_r(\vecx,t)
\chi(\nu)$, and then defining the integrated quantity
\begin{equation}
\label{eq:grey_radiation_energy}
   E_r(\vecx,t) \equiv \int_{\nu_0}^{\infty}
   E_{\nu}(\nu,\vecx,t)\,\mathrm d\nu \  = \ 
   \tilde{E}_r(\vecx,t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
\end{equation}
The quantity $D$ in Equation (\ref{eq:fld_radiation}) is the Larsen
square-root flux-limiter \citep[see][]{Morel2000}, $\kappa$ is  the
total opacity, $\eta$ is the field of radiation sources, $\Gamma$ is
the radiation-induced photo-heating rate, and $\Lambda$ is the
chemical cooling rate.
%, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

%\subsubsection{Thermal conduction}

@@

The grid hierarchy begins with the root grid, which covers the entire
domain of interest with a coarse uniform Cartesian grid. Then, as
the solution evolves and interesting regions start to develop, finer grids are
placed within these coarse regions.
% (we use the notation `below' to refer to
% finer grids and `above' for coarser grids). 
We restrict the ratio between cell sizes of parent and child grids to be an integer, typically 2
or 4, and refer to a level as all grids with the same cell size.  For simplicity, the edges of subgrids must
coincide with the cell edge of its immediate parent (coarser)
grid. Additionally, the hierarchy can be initialized with one or more
static grids if a higher initial resolution is required.

@@

Since interesting regions on the grid may move, the hierarchy must
adapt itself. This happens whenever a level has caught up to its
parent level, by entirely rebuilding the grids on that level and at
finer resolutions. Rebuilding is achieved by applying the grid
refinement criteria to the grids on that level and flagging zones that
require extra grids.  These criteria depend on the physical problem
being simulated; see Section~\ref{sec:refinement_criteria} for more
details.  Once a grid has a set of flagged cells, we apply a
machine-vision based algorithm \citep{Berger91} to find edges and
determine a good placement of subgrids.
%  These subgrids must not overlap one another, must cover all flagged
% cells and their neighbouring cells, and be above a preset efficiency
% threshold, where the efficiency of a cell is defined as the ratio of
% flagged cells to total cells. 
Once these new subgrids have been identified, the solution from the
next coarser grid is interpolated in order to initialize the values on
the new grids.  Finally, any overlap between new subgrids and old ones
is identified, and the prior solution within the regions of overlap is
copied to the new subgrids. This procedure is then repeated on the new
grids and in this way, the entire hierarchy (from the original level
examined and continuing on to all finer levels) is rebuilt.

@@

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order accurate version of Godunov's method for ideal
gas dynamics with the spatially third-order accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
%\begin{enumerate}
% \item 
(i) Construct monotonic parabolic interpolation of cell average data,
for each fluid quantity;
% \item 
(ii) Compute interface states by averaging the parabola over the
domain of dependence for each interface;
% \item 
(iii) Use interface data to solve the Riemann problem;
 %\item 
(iv) Difference the interface fluxes to update the cell average
quantities.
%\end{enumerate}

@@

Although the particles are fixed in mass once initialized (with the
exception of star particles, which can lose mass in the feedback
process), we can create them with any initial set of masses and
positions.  For example, in many cosmological simulations static
subgrids are included from the beginning in order to improve the
initial spatial and baryonic mass resolution, and these subgrids are
populated with lower-mass particles to correspondingly improve the
collisionless mass resolution.
%% One particle per initial grid point seems to provide approximately
%% equal sampling between the dark matter and gas. \mqk{This
%% contradicts the conclusions of O'Shea et al. 2005 that an initial
%% grid twice as fine as the mean particle separation is necessary to
%% properly capture the abundance of low mass halos.}

@@

\enzo\ includes the capability of following up to 12 particle species
using a non-equilibrium chemical network.  The species can be turned
on in sets, with the simplest model including just atomic hydrogen and
helium (H, H$^+$, He, He$^+$, He$^{++}$, e$^-$), and more complete
models adding first species important for gas-phase molecular hydrogen
formation (H$^-$, H$_2$ and H$_2^+$), and then HD formation (HD, D,
D$^+$).
%(We note that a reduced version
%of the chemistry, used only with the flux-limited diffusion radiation
%transport, includes only hydrogen species.)
The cooling and heating due to these species is included (see the next
section). The solution of the rate equations is carried out using one
Jacobi iteration of an implicit Euler time discretization to ensure
stability. To ensure accuracy the rate equations are sub-cycled
within one hydrodynamic timestep, subject to the constraint that the
electron and neutral fractions do not change by more than 10\% in one
sub-cycle.

>>>>

@@

\begin{eqnarray} 
  \frac{\partial \rho}{\partial t} 
  + \frac{1}{a} \div (\rho \vecv) & = & 0,
  \label{eq:mass} \\
%
  \frac{\partial \rho \vecv}{\partial t}  
  + \frac{1}{a} \div \left(\rho \vecv \vecv + \myvec{I}p^* - 
    \frac{\vecB \vecB}{a} \right) & = &
  - \frac{\dot{a}}{a} \rho \vecv - \frac{1}{a} \rho \grad \phi,
  \label{eq:momentum} \\
%
  \frac{\partial E}{\partial t} 
  + \frac{1}{a} \div \left[ (E + p^*) \vecv - 
    \frac{1}{a} \vecB (\vecB \cdot \vecv) \right] & = &
  - \frac{\dot{a}}{a} \left( 2E - \frac{B^2}{2a} \right) - 
  \frac{\rho}{a} \vecv \cdot \grad \phi 
  - \Lambda + \Gamma + \frac{1}{a^2} \div \fcond,
  \label{eq:total_energy}  \\
%
  \frac{\partial \vecB}{\partial t} - 
  \frac{1}{a}  \grad \times (\vecv \times \vecB) & = & 
  %-\frac{\dot{a}}{2 a} \vecB. 
  0
  \label{eq:induction}
\end{eqnarray}

In these equations, $E$, $\rho$, $\vecv$, and $\vecB$ are the comoving
total fluid energy density, comoving gas density, peculiar
velocity, and comoving magnetic field strength, respectively. The
matrix 
$\myvec{I}$ is the identity matrix, and $a$ is the cosmological
expansion parameter (discussed in more detail below).  The first
equation represents conservation of mass, the second conservation of
momentum, and the third conservation of total (kinetic plus thermal
plus magnetic) fluid energy.  They are respectively, the first, second, and third
moments of the Boltzmann equation.  The fourth equation is the
magnetic induction equation.  Terms representing radiative cooling
($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the
energy equation (\ref{eq:total_energy}), as does the flux due to
thermal heat conduction ($\fcond$).

@@

\begin{eqnarray}
\vecx & = & \vecx' / a, \\
\vecv & = & a \hspace{0.5mm} \dot{\vecx} = 
              \vecvp - \dot{a}\vecx, \\
\rho  & = & a^3 \rhop,   \\
p     & = & a^3 \pp, \\
E     & = & a^3 \left(\Ep - 
            \dot{a} \, \vecx \cdot \vecvp - 
            \frac{1}{2} \dot{a}^2 \vecx^2\right), \\
\phi  & = & \phip + \frac{1}{2} \, a \ddot{a} \, \vecx^2, \\
\vecB & = & a^2 \vecB^\prime,
\end{eqnarray}
where primes indicate quantities in proper coordinates (fixed frame).
We note that the definition of comoving $\vecB$ used here is such
that a uniform field is constant in a homogeneous expanding universe
and is also the quantity used in the MHD-CT solver described
below; however, it is not universal,
and is slightly different from that used both in \citet{Li08a}
and in the MHD-Dedner solver (see section~\ref{sec.num.hydro-muscl}
for more details).
The expansion parameter $a \equiv 1/(1 + z)$ follows the expansion of
a smooth, homogeneous background, where $z$, the redshift, is a
function only of $t$.  All spatial derivatives are determined with
respect to the comoving position $\vecx$, which removes the universal
expansion from the coordinate system. The evolution of $a(t)$ is
governed by the second Friedmann equation for the expansion of a
spatially homogeneous and isotropic universe
%
\begin{eqnarray}
\frac{\ddot{a}}{a} & = & 
      - \frac{4 \pi G }{3 a^3 } (\rho_0 
      + 3p_0/c^2) 
      + \Lambda_c c^2 / 3 .
      \label{eq:expansion} 
\end{eqnarray}
%
Here $\rho_0$ is the mean comoving mass density (including both
baryonic and dark matter), $p_0$ is the comoving background pressure
contribution, and $\Lambda_c$ is the cosmological constant.
This system of equations is limited to the non-relativistic regime and
assumes that curvature effects are not important --- both assumptions
are reasonable as long as the size of the simulated region is small
compared to the radius of curvature and the Hubble length $c/H$, where
$c$ is the speed of light and $H = \dot{a}/a$ is the Hubble constant.

@@

@@

@@

As an alternative to the ray-casting strategy, \enzo\ can also solve
the angle-averaged radiative transfer equation through a
flux-limited diffusion approximation, with couplings to both the gas
energy and chemical number densities. The equations solved are
\begin{eqnarray}
  \label{eq:fld_radiation}
  \frac{\partial E_r}{\partial t} + \frac1a \div\left(E_r \vecv \right) &=& 
  \div\left(D\grad E_r\right) -
  \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
  \label{eq:fld_heating}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c + \Gamma - \Lambda,
\end{eqnarray}
where $E_r = E_r(\vecx,t)$ is a grey radiation energy density and
$e_c$ is the internal energy correction due to photo-heating and
chemical cooling.
Here, we define $E_r$ through first assuming a fixed frequency
spectrum, i.e.  $E_{\nu}(\nu,\vecx,t) = \tilde{E}_r(\vecx,t)
\chi(\nu)$, and then defining the integrated quantity
\begin{equation}
\label{eq:grey_radiation_energy}
   E_r(\vecx,t) \equiv \int_{\nu_0}^{\infty}
   E_{\nu}(\nu,\vecx,t)\,\mathrm d\nu \  = \ 
   \tilde{E}_r(\vecx,t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
\end{equation}
The quantity $D$ in Equation (\ref{eq:fld_radiation}) is the Larsen
square-root flux-limiter \citep[see][]{Morel2000}, $\kappa$ is  the
total opacity, $\eta$ is the field of radiation sources, $\Gamma$ is
the radiation-induced photo-heating rate, and $\Lambda$ is the
chemical cooling rate.

@@

The grid hierarchy begins with the root grid, which covers the entire
domain of interest with a coarse uniform Cartesian grid. Then, as
the solution evolves and interesting regions start to develop, finer grids are
placed within these coarse regions.
We restrict the ratio between cell sizes of parent and child grids to be an integer, typically 2
or 4, and refer to a level as all grids with the same cell size.  For simplicity, the edges of subgrids must
coincide with the cell edge of its immediate parent (coarser)
grid. Additionally, the hierarchy can be initialized with one or more
static grids if a higher initial resolution is required.

@@

Since interesting regions on the grid may move, the hierarchy must
adapt itself. This happens whenever a level has caught up to its
parent level, by entirely rebuilding the grids on that level and at
finer resolutions. Rebuilding is achieved by applying the grid
refinement criteria to the grids on that level and flagging zones that
require extra grids.  These criteria depend on the physical problem
being simulated; see Section~\ref{sec:refinement_criteria} for more
details.  Once a grid has a set of flagged cells, we apply a
machine-vision based algorithm \citep{Berger91} to find edges and
determine a good placement of subgrids.
Once these new subgrids have been identified, the solution from the
next coarser grid is interpolated in order to initialize the values on
the new grids.  Finally, any overlap between new subgrids and old ones
is identified, and the prior solution within the regions of overlap is
copied to the new subgrids. This procedure is then repeated on the new
grids and in this way, the entire hierarchy (from the original level
examined and continuing on to all finer levels) is rebuilt.

@@

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order accurate version of Godunov's method for ideal
gas dynamics with the spatially third-order accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
(i) Construct monotonic parabolic interpolation of cell average data,
for each fluid quantity;
(ii) Compute interface states by averaging the parabola over the
domain of dependence for each interface;
(iii) Use interface data to solve the Riemann problem;
(iv) Difference the interface fluxes to update the cell average
quantities.

@@

Although the particles are fixed in mass once initialized (with the
exception of star particles, which can lose mass in the feedback
process), we can create them with any initial set of masses and
positions.  For example, in many cosmological simulations static
subgrids are included from the beginning in order to improve the
initial spatial and baryonic mass resolution, and these subgrids are
populated with lower-mass particles to correspondingly improve the
collisionless mass resolution.

@@

\enzo\ includes the capability of following up to 12 particle species
using a non-equilibrium chemical network.  The species can be turned
on in sets, with the simplest model including just atomic hydrogen and
helium (H, H$^+$, He, He$^+$, He$^{++}$, e$^-$), and more complete
models adding first species important for gas-phase molecular hydrogen
formation (H$^-$, H$_2$ and H$_2^+$), and then HD formation (HD, D,
D$^+$).
The cooling and heating due to these species is included (see the next
section). The solution of the rate equations is carried out using one
Jacobi iteration of an implicit Euler time discretization to ensure
stability. To ensure accuracy the rate equations are sub-cycled
within one hydrodynamic timestep, subject to the constraint that the
electron and neutral fractions do not change by more than 10\% in one
sub-cycle.
